# Reproducible Research: Peer Assessment 1
Last modified on 2014-06-15

Report generated on `r date()`
 
-------------------------
## Loading and preprocessing the data
```{r}
setwd("~/GitHub/RepData_PeerAssessment1")
activity <- read.csv("~/GitHub/RepData_PeerAssessment1/activity.csv")
activity$date <- as.Date(activity$date) #set date format correctly

```

----------------------------
## What is mean total number of steps taken per day?
```{r}
steps_by_day <- tapply(activity$steps, activity$date, sum)
hist(steps_by_day, main= "Histogram of Total Steps per Day", xlab="Steps per Day")
```
**The mean number of steps per day (excluding the missing data) was `r format(mean(steps_by_day, na.rm=T), scientific=F, nsmall=2)`, and the median number of steps per day was `r median(steps_by_day, na.rm=T)`.**

------------------------------
## What is the average daily activity pattern?
```{r}
steps_by_interval <- tapply(activity$steps, activity$interval, mean, na.rm=T)
plot(x=names(steps_by_interval), y=steps_by_interval, type='l', main = "Average # Steps per Interval", xlab= "Interval", ylab= "Average # Steps")
```
**The interval with the greatest number of steps (averaged across all days) is interval `r names(steps_by_interval[which(steps_by_interval == max(steps_by_interval))])`, with a mean of `r steps_by_interval[which(steps_by_interval == max(steps_by_interval))]` steps during that interval.**


------------------------------
## Imputing missing values
There are `r sum(is.na(activity$steps))` missing _steps_ values (as NA) in the _activity_ dataset. We will impute these missing values using the mean number of steps for the associated interval across the entire dataset.
```{r}
imp_activity <- activity
imp_activity$imp_steps <- NA #preset the column to hold imputed values to NA

# Imputation loop below: 
# substitutes interval mean steps where steps is missing
for(i in 1:nrow(imp_activity)){
  if(is.na(imp_activity$steps[i])){
    imp_activity$imp_steps[i] <- steps_by_interval[names(steps_by_interval) == imp_activity$interval[i]] 
    } else {
      imp_activity$imp_steps[i] <- imp_activity$steps[i]
      }
  }

imp_steps_by_day <- tapply(imp_activity$imp_steps, imp_activity$date, sum)
hist(imp_steps_by_day, main= "Histogram of Total Steps per Day", xlab="Steps per Day")
```

**The mean number of total steps per day (including the newly-imputed data) was `r format(mean(imp_steps_by_day, na.rm=T), scientific=F, nsmall=2)`, and the median number of steps per day was `r format(median(imp_steps_by_day, na.rm=T), scientific=F)`. Note that the mean total steps per day did not change under imputation, and the median increased by one step.**



------------------------------
## Are there differences in activity patterns between weekdays and weekends?
```{r}
# To create the weekday/weekend flag
imp_activity$day_of_week <- weekdays(imp_activity$date, abbreviate=T)
imp_activity$day_type <- ifelse(imp_activity$day_of_week %in% c("Sat","Sun"),"Weekend", "Weekday")

# Average steps per interval across weekday/weekend split
weekday_split <- split(imp_activity, imp_activity$day_type)

weekend_steps_by_interval <- data.frame(tapply(weekday_split$Weekend$imp_steps, weekday_split$Weekend$interval, mean))
weekend_steps_by_interval$day_type <- "Weekend"
weekend_steps_by_interval$interval <- rownames(weekend_steps_by_interval)
colnames(weekend_steps_by_interval) <- c("steps","day_type","interval")

weekday_steps_by_interval <- data.frame(tapply(weekday_split$Weekday$imp_steps, weekday_split$Weekday$interval, mean))
weekday_steps_by_interval$day_type <- "Weekday"
weekday_steps_by_interval$interval <- rownames(weekday_steps_by_interval)
colnames(weekday_steps_by_interval) <- c("steps","day_type","interval")

par(mfrow=c(2,1))
plot(x=weekday_steps_by_interval$interval, y=weekday_steps_by_interval$steps, type='l', main = "Weekday", xlab= "Interval", ylab= "Average Steps", las=1)
plot(x=weekend_steps_by_interval$interval, y=weekend_steps_by_interval$steps, type='l', main = "Weekend", xlab= "Interval", ylab= "Average Steps", las=1)

```



=======================================
